Joint estimation of ancestral states and phylogenetic model parameters is inconsistent

### 2017-11-30 appendix.tex
minor edits; removing redundancy

Let
$$
L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t)) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the limit of infinite data,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log-likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
\end{equation}

@@

\paragraph{Example}
\begin{figure}
    \centering
    \includegraphics[width=.45\textwidth]{unrooted_farris_tree}
    \caption{Four taxon ``Farris zone'' tree.}
\label{fig:four-taxa-tree}
\end{figure}

@@

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
    \hline
     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$          &$(1+x)^2   (1-w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1+x)^2   (1+w)(1-y)^2$          &$(1+x)^2   (1-w)(1+y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
    \hline
    \hline
    &$\{1\}$                             &$\{1,2\}$  \\
    \hline
     $\emptyset$   &$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
All values multiplied by $1/32$.}
\label{tab:farris_likelihoods}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_2,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
    \hline
     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1-x)^2   (1-w)(1+y)(1-y)$\\
     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1+x)^2   (1-w)(1+y)^2$\\
     $\{1,3\}$     &$(1+x)^2   (1+w)(1-y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1-x)^2   (1-w)(1+y)^2$\\
     $\{1,2,3\}$   &$(1+x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1+y)^2$\\
    \hline
    \hline
    &$\{1\}$                             &$\{1,2\}$  \\
    \hline
     $\emptyset$   &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       &$(1+x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       &$(1+x)(1-x)(1-w)(1+y)^2$           &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     &$(1-x)^2   (1-w)(1-y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1-x)^2   (1+w)(1+y)^2$\\
     $\{2,3\}$     &$(1+x)^2   (1-w)(1-y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1+x)(1-x)(1-w)(1-y)^2$           &$(1-x)^2   (1+w)(1+y)(1-y)$\\
\hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and internal states of Felsenstein topology.
All values multiplied by $1/32$.}
\label{tab:fels_likelihoods}
\end{table}

@@

Consider the Farris tree with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
We now show that, in the case of the Farris tree topology, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
Using the Hadamard transform approach, we calculate the generating probabilities on the Farris tree topology.
For site split $\emptyset$,
\begin{align*}
    P(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities will differ only in the signs of each term.
For example, for site split $\{1\}$ we have
\begin{align*}
    P(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    P(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
\begin{align*}
    P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

@@

Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
The analogous statement holds for $y_1$ and $y_2$.
The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

We take various safeguards to ensure convergence and stability of the procedure.
Since this method involves randomness, we take precautions to not evaluate the likelihood near or outside of boundary conditions, i.e., when $x^*$ or $y^*$ are either $0$ or $1$.
As is apparent from Table~\ref{tab:sitepatprob}, if $0 < x^*, y^* < 1$ and we try to evaluate the candidate $\hat{x}=\hat{y}=\hat{w}=1$ we have a likelihood of exactly zero.
Worse still, in Table~\ref{tab:likelihoods} for the Farris topology, in the cases of $\{2,\}, \{1,2\}, \{2,3\},$ and $\{1,2,3\}$, any attempt to evaluate $\hat{y}=1$ will similarly yield a zero likelihood.
In these cases of zero likelihood, methods that evaluate many candidate parameter values can fail to converge when near these boundaries.
We sidestep these computational issues by restricting ourselves to the region $x^*,y^*\in [10^{-2},1-10^{-2}]^2$.

>>>>

Let
$$
L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t)) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, as $\nCols\rightarrow\infty$,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log-likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
\end{equation}

@@

\paragraph{Example}
\begin{figure}
    \centering
    \includegraphics[width=.45\textwidth]{unrooted_farris_tree}
    \caption{Four taxon ``Farris zone'' topology.}
\label{fig:four-taxa-tree}
\end{figure}

@@

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
    \hline
     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$          &$(1+x)^2   (1-w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1+x)^2   (1+w)(1-y)^2$          &$(1+x)^2   (1-w)(1+y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
    \hline
    \hline
    &$\{1\}$                             &$\{1,2\}$  \\
    \hline
     $\emptyset$   &$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and internal states of Farris tree.
All values multiplied by $1/32$.}
\label{tab:farris_likelihoods}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_2,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
    \hline
     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1-x)^2   (1-w)(1+y)(1-y)$\\
     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1+x)^2   (1-w)(1+y)^2$\\
     $\{1,3\}$     &$(1+x)^2   (1+w)(1-y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1-x)^2   (1-w)(1+y)^2$\\
     $\{1,2,3\}$   &$(1+x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1+y)^2$\\
    \hline
    \hline
    &$\{1\}$                             &$\{1,2\}$  \\
    \hline
     $\emptyset$   &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       &$(1+x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       &$(1+x)(1-x)(1-w)(1+y)^2$           &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     &$(1-x)^2   (1-w)(1-y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1-x)^2   (1+w)(1+y)^2$\\
     $\{2,3\}$     &$(1+x)^2   (1-w)(1-y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1+x)(1-x)(1-w)(1-y)^2$           &$(1-x)^2   (1+w)(1+y)(1-y)$\\
\hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and internal states of Felsenstein tree.
All values multiplied by $1/32$.}
\label{tab:fels_likelihoods}
\end{table}

@@

Consider the Farris tree with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
We now show that, in the case of the Farris tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
Using the Hadamard transform approach, we calculate the generating probabilities on the Farris tree.
For site split $\emptyset$,
\begin{align*}
    P(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities will differ only in the signs of each term.
For example, for site split $\{1\}$ we have
\begin{align*}
    P(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    P(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
\begin{align*}
    P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

@@

Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
The analogous statement holds for $y_1$ and $y_2$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

We take various safeguards to ensure convergence and stability of the procedure.
Since this method involves randomness, we take precautions to not evaluate the likelihood near or outside of boundary conditions, i.e., when $x^*$ or $y^*$ are either $0$ or $1$.
As is apparent from Table~\ref{tab:sitepatprob}, if $0 < x^*, y^* < 1$ and we try to evaluate the candidate $\hat{x}=\hat{y}=\hat{w}=1$ we have a likelihood of exactly zero.
Worse still, in Table~\ref{tab:likelihoods} for the Farris tree, in the cases of $\{2,\}, \{1,2\}, \{2,3\},$ and $\{1,2,3\}$, any attempt to evaluate $\hat{y}=1$ will similarly yield a zero likelihood.
In these cases of zero likelihood, methods that evaluate many candidate parameter values can fail to converge when near these boundaries.
We sidestep these computational issues by restricting ourselves to the region $x^*,y^*\in [10^{-2},1-10^{-2}]^2$.
